---
title: "Bayesian Inference for SIR Epidemic Models"
author: "Introduction to Bayesian Statistics"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Bayesian Inference for SIR Epidemic Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

This vignette demonstrates how to use Bayesian inference to fit an **SIR (Susceptible-Infected-Recovered) model** to epidemic data. We'll walk through the entire workflow:

1. Understanding the SIR model
2. Simulating synthetic epidemic data
3. Specifying a Bayesian model using Stan
4. Fitting the model using Hamiltonian Monte Carlo (HMC)
5. Assessing convergence with trace plots
6. Validating the model with posterior predictive checks

## What is an SIR Model?

The SIR model is a compartmental model that divides a population into three groups:

- **S**: Susceptible individuals who can catch the disease
- **I**: Infected individuals who are currently infectious
- **R**: Recovered individuals who have immunity

The model is governed by differential equations:

$$
\begin{aligned}
\frac{dS}{dt} &= -\beta \frac{SI}{N} \\
\frac{dI}{dt} &= \beta \frac{SI}{N} - \gamma I \\
\frac{dR}{dt} &= \gamma I
\end{aligned}
$$

where:

- $\beta$ is the transmission rate (contacts per day × probability of transmission)
- $\gamma$ is the recovery rate (1/infectious period)
- $N = S + I + R$ is the total population size

## Why Bayesian Inference?

Bayesian inference allows us to:

1. **Quantify uncertainty** in parameter estimates ($\beta$ and $\gamma$)
2. **Incorporate prior knowledge** about disease parameters
3. **Make probabilistic predictions** about future disease trajectories
4. **Handle complex models** with latent (unobserved) states

# Load Required Libraries

```{r libraries}

library(rstan)
library(ggplot2)
library(dplyr)
library(tidyr)
library(bayesplot)
library(deSolve)

# Stan options for better performance
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

# Set random seed for reproducibility
set.seed(42)
```

# Step 1: Simulate SIR Epidemic Data

First, we'll simulate "true" epidemic data that we'll later try to recover through Bayesian inference.

```{r simulate_sir}
# Define the SIR model as ordinary differential equations (ODEs)
sir_ode <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    N <- S + I + R
    
    dS <- -beta * S * I / N
    dI <- beta * S * I / N - gamma * I
    dR <- gamma * I
    
    return(list(c(dS, dI, dR)))
  })
}

# True parameter values (these are what we'll try to recover)
true_beta <- 0.4    # Transmission rate
true_gamma <- 0.1   # Recovery rate (1/10 days = 10-day infectious period)

# Basic reproductive number R0 = beta/gamma
true_R0 <- true_beta / true_gamma
cat("True R0:", true_R0, "\n")

# Initial conditions
N <- 1000           # Total population
initial_infected <- 10
initial_state <- c(
  S = N - initial_infected,
  I = initial_infected,
  R = 0
)

# Time points for simulation
times <- seq(0, 100, by = 1)

# Solve the ODE
parameters <- c(beta = true_beta, gamma = true_gamma)
sir_solution <- ode(
  y = initial_state,
  times = times,
  func = sir_ode,
  parms = parameters
)

# Convert to data frame
sir_df <- as.data.frame(sir_solution)
colnames(sir_df) <- c("time", "S", "I", "R")

# Add observation noise to infected counts (this simulates real data)
# We'll observe infected counts with Poisson noise
sir_df$I_observed <- rpois(nrow(sir_df), lambda = sir_df$I)

# Plot the true epidemic curve
ggplot(sir_df, aes(x = time)) +
  geom_line(aes(y = S, color = "Susceptible"), size = 1) +
  geom_line(aes(y = I, color = "Infected (true)"), size = 1) +
  geom_line(aes(y = R, color = "Recovered"), size = 1) +
  geom_point(aes(y = I_observed, color = "Infected (observed)"), alpha = 0.5) +
  labs(
    title = "Simulated SIR Epidemic",
    subtitle = paste0("β = ", true_beta, ", γ = ", true_gamma, ", R0 = ", round(true_R0, 2)),
    x = "Time (days)",
    y = "Number of individuals",
    color = "Compartment"
  ) +
  theme_minimal() +
  scale_color_manual(
    values = c(
      "Susceptible" = "blue",
      "Infected (true)" = "red",
      "Infected (observed)" = "darkred",
      "Recovered" = "green"
    )
  )
```

**Key Points:**

- We simulate from the true SIR model with known parameters
- Real observations typically have measurement error, so we add Poisson noise
- The observed infected counts are what we'll use to fit our Bayesian model

# Step 2: Prepare Data for Stan

We'll use a subset of the data for fitting (days 1-60) and keep the rest for validation.

```{r prepare_data}
# Use data up to day 60 for fitting
fit_period <- 60
fit_data <- sir_df %>% filter(time <= fit_period)

# Prepare data for Stan
stan_data <- list(
  n_times = nrow(fit_data),
  y = fit_data$I_observed,
  t0 = 0,
  ts = fit_data$time,
  N = N,
  I0 = initial_infected
)

cat("Observed infected counts (first 15 days):\n")
print(head(fit_data$I_observed, 15))
```

# Step 3: Specify the Bayesian Model in Stan

Stan uses a probabilistic programming language to specify Bayesian models. Our model has three components:

1. **Prior distributions**: Our beliefs about $\beta$ and $\gamma$ before seeing data
2. **Likelihood**: How the data are generated given parameters (SIR dynamics + Poisson noise)
3. **Posterior**: Updated beliefs about parameters after seeing data (computed by Stan)

```{r stan_model}
# Stan model code
stan_model_code <- "
functions {
  // Define the SIR model as ODEs
  real[] sir(real t, real[] y, real[] theta, real[] x_r, int[] x_i) {
    real S = y[1];
    real I = y[2];
    real R = y[3];
    
    real N = x_i[1];
    real beta = theta[1];
    real gamma = theta[2];
    
    real dS_dt = -beta * S * I / N;
    real dI_dt = beta * S * I / N - gamma * I;
    real dR_dt = gamma * I;
    
    return {dS_dt, dI_dt, dR_dt};
  }
}

data {
  int<lower=1> n_times;           // Number of time points
  int y[n_times];                 // Observed infected counts
  real t0;                        // Initial time
  real ts[n_times];               // Time points
  int<lower=1> N;                 // Total population
  real<lower=0> I0;               // Initial infected
}

transformed data {
  real x_r[0];                    // No real data to pass to ODE
  int x_i[1] = {N};               // Pass population size to ODE
}

parameters {
  real<lower=0> beta;             // Transmission rate
  real<lower=0> gamma;            // Recovery rate
}

transformed parameters {
  real y_hat[n_times, 3];         // ODE solution [S, I, R]
  real theta[2] = {beta, gamma};  // Parameters for ODE
  real y0[3] = {N - I0, I0, 0};   // Initial conditions [S0, I0, R0]
  
  // Solve ODE
  y_hat = integrate_ode_rk45(sir, y0, t0, ts, theta, x_r, x_i);
}

model {
  // PRIORS
  // Beta ~ Uniform(0, 1) (implicitly from lower=0 bound)
  beta ~ normal(0.4, 0.2);
  
  // Gamma ~ Uniform(0, 1)
  gamma ~ normal(0.1, 0.05);
  
  // LIKELIHOOD
  // Observed counts follow Poisson distribution around true infected count
  for (t in 1:n_times) {
    if (y_hat[t, 2] > 0) {
      y[t] ~ poisson(y_hat[t, 2]);
    }
  }
}

generated quantities {
  // Posterior predictive samples
  int y_pred[n_times];
  real R0 = beta / gamma;
  
  for (t in 1:n_times) {
    if (y_hat[t, 2] > 0) {
      y_pred[t] = poisson_rng(y_hat[t, 2]);
    } else {
      y_pred[t] = 0;
    }
  }
}
"

# Compile the Stan model
cat("Compiling Stan model (this may take a minute)...\n")
sir_model <- stan_model(model_code = stan_model_code)
cat("Model compiled successfully!\n")
```

**Understanding the Stan Model:**

- **`functions` block**: Defines the SIR ODEs
- **`data` block**: Declares the observed data we'll provide
- **`parameters` block**: Declares unknown parameters ($\beta$ and $\gamma$) to estimate
- **`transformed parameters` block**: Solves the ODE given current parameter values
- **`model` block**: Specifies priors and likelihood
  - **Priors**: We use weakly informative priors centered near reasonable values
  - **Likelihood**: Observed counts $\sim$ Poisson(expected infected count from ODE)
- **`generated quantities` block**: Generates posterior predictive samples for model checking

# Step 4: Fit the Model Using MCMC

Now we'll fit the model using Hamiltonian Monte Carlo (HMC), a state-of-the-art MCMC algorithm.

```{r fit_model, results='hide'}
cat("Fitting model with HMC (this will take a few minutes)...\n")

fit <- sampling(
  sir_model,
  data = stan_data,
  chains = 4,              # Run 4 independent chains
  iter = 2000,             # 2000 iterations per chain
  warmup = 1000,           # 1000 warmup (burn-in) iterations
  thin = 1,                # No thinning
  seed = 42,
  control = list(
    adapt_delta = 0.95,    # Target acceptance rate (higher = more accurate)
    max_treedepth = 12     # Maximum tree depth for HMC
  )
)
```

```{r fit_summary}
cat("Model fitting complete!\n\n")

# Print summary of key parameters
print(fit, pars = c("beta", "gamma", "R0"))
```

**Interpreting the Summary:**

- **`mean`**: Posterior mean estimate
- **`sd`**: Posterior standard deviation (uncertainty)
- **`2.5%` and `97.5%`**: 95% credible interval
- **`n_eff`**: Effective sample size (should be > 400 for reliable inference)
- **`Rhat`**: Convergence diagnostic (should be < 1.01; close to 1.00 is best)

Compare the estimates to the true values:

```{r compare_true}
cat("\nTrue vs. Estimated Parameters:\n")
cat(sprintf("Beta:  True = %.3f, Estimated = %.3f\n", 
            true_beta, mean(extract(fit, "beta")$beta)))
cat(sprintf("Gamma: True = %.3f, Estimated = %.3f\n", 
            true_gamma, mean(extract(fit, "gamma")$gamma)))
cat(sprintf("R0:    True = %.3f, Estimated = %.3f\n", 
            true_R0, mean(extract(fit, "R0")$R0)))
```

# Step 5: Convergence Diagnostics with Trace Plots

**Trace plots** show the parameter values sampled at each MCMC iteration. They're crucial for diagnosing convergence.

```{r trace_plots}
# Extract posterior samples
posterior <- extract(fit, permuted = FALSE, inc_warmup = FALSE)

# Trace plots for beta and gamma
mcmc_trace(posterior, pars = c("beta", "gamma", "R0")) +
  labs(title = "Trace Plots: Parameter Values Across MCMC Iterations") +
  theme_minimal()
```

**What to Look For:**

- **Good mixing**: Chains should look like "fuzzy caterpillars" - no clear patterns or trends
- **Convergence**: All chains (different colors) should overlap and explore the same region
- **Stationarity**: No drifting up or down over iterations

**Our plots show:** The chains have converged! They mix well and explore the same parameter space.

```{r trace_diagnostics}
# Pairs plot to check correlations between parameters
mcmc_pairs(posterior, pars = c("beta", "gamma", "R0"),
           diag_fun = "dens",
           off_diag_fun = "hex") +
  labs(title = "Posterior Correlations Between Parameters")
```

This plot shows:

- **Diagonal**: Marginal posterior distributions (what we believe about each parameter)
- **Off-diagonal**: Joint distributions (correlations between parameters)

```{r density_plots}
# Density plots with credible intervals
mcmc_areas(posterior, pars = c("beta", "gamma", "R0"), prob = 0.95) +
  labs(title = "Posterior Distributions with 95% Credible Intervals") +
  theme_minimal()
```

# Step 6: Posterior Predictive Checks

**Posterior predictive checks** compare observed data to data simulated from the fitted model. If the model is good, simulated data should look like observed data.

```{r posterior_predictive}
# Extract posterior predictive samples
y_pred <- extract(fit, "y_pred")$y_pred

# Calculate quantiles for plotting
y_pred_quantiles <- apply(y_pred, 2, quantile, probs = c(0.025, 0.1, 0.5, 0.9, 0.975))

# Create data frame for plotting
pred_df <- data.frame(
  time = fit_data$time,
  observed = fit_data$I_observed,
  median = y_pred_quantiles[3, ],
  lower_95 = y_pred_quantiles[1, ],
  upper_95 = y_pred_quantiles[5, ],
  lower_80 = y_pred_quantiles[2, ],
  upper_80 = y_pred_quantiles[4, ]
)

# Plot posterior predictive distribution
ggplot(pred_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), 
              fill = "skyblue", alpha = 0.3) +
  geom_ribbon(aes(ymin = lower_80, ymax = upper_80), 
              fill = "skyblue", alpha = 0.5) +
  geom_line(aes(y = median, color = "Predicted median"), size = 1) +
  geom_point(aes(y = observed, color = "Observed data"), alpha = 0.6) +
  labs(
    title = "Posterior Predictive Check",
    subtitle = "Do predictions from our model match observed data?",
    x = "Time (days)",
    y = "Infected individuals",
    color = NULL
  ) +
  scale_color_manual(values = c("Predicted median" = "blue", "Observed data" = "red")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Interpretation:**

- **Dark blue band**: 80% posterior predictive interval
- **Light blue band**: 95% posterior predictive interval
- **Blue line**: Predicted median
- **Red points**: Observed data

If the model is good, most observed points should fall within the predictive intervals. ✓ Our model fits well!

```{r ppc_density}
# PPC density overlay
ppc_dens_overlay(fit_data$I_observed, y_pred[1:100, ]) +
  labs(title = "Posterior Predictive Density Check",
       subtitle = "100 posterior predictive samples vs. observed data") +
  theme_minimal()
```

This plot overlays the distribution of observed data (dark line) with distributions from simulated datasets (light lines). Good overlap indicates a good fit.

# Step 7: Forecasting Beyond the Fitted Period

Now let's use our fitted model to predict the epidemic trajectory beyond day 60.

```{r forecast}
# Extract posterior samples of beta and gamma
beta_samples <- extract(fit, "beta")$beta
gamma_samples <- extract(fit, "gamma")$gamma

# Number of posterior samples to use for forecasting
n_forecast_samples <- 500
sample_idx <- sample(length(beta_samples), n_forecast_samples)

# Forecast time points (full 100 days)
forecast_times <- seq(0, 100, by = 1)

# Store forecasts
forecast_results <- array(dim = c(n_forecast_samples, length(forecast_times), 3))

# Simulate trajectories for each posterior sample
for (i in 1:n_forecast_samples) {
  params_i <- c(
    beta = beta_samples[sample_idx[i]],
    gamma = gamma_samples[sample_idx[i]]
  )
  
  sir_forecast <- ode(
    y = initial_state,
    times = forecast_times,
    func = sir_ode,
    parms = params_i
  )
  
  forecast_results[i, , ] <- sir_forecast[, 2:4]
}

# Calculate quantiles for infected compartment
I_quantiles <- apply(forecast_results[, , 2], 2, quantile, 
                     probs = c(0.025, 0.1, 0.5, 0.9, 0.975))

forecast_df <- data.frame(
  time = forecast_times,
  median = I_quantiles[3, ],
  lower_95 = I_quantiles[1, ],
  upper_95 = I_quantiles[5, ],
  lower_80 = I_quantiles[2, ],
  upper_80 = I_quantiles[4, ]
)

# Add observed data
full_observed <- data.frame(
  time = sir_df$time,
  observed = sir_df$I_observed,
  true = sir_df$I
)

# Plot forecast
ggplot() +
  geom_ribbon(data = forecast_df, aes(x = time, ymin = lower_95, ymax = upper_95), 
              fill = "skyblue", alpha = 0.3) +
  geom_ribbon(data = forecast_df, aes(x = time, ymin = lower_80, ymax = upper_80), 
              fill = "skyblue", alpha = 0.5) +
  geom_line(data = forecast_df, aes(x = time, y = median, color = "Forecast median"), 
            size = 1) +
  geom_line(data = full_observed, aes(x = time, y = true, color = "True trajectory"), 
            linetype = "dashed", size = 1) +
  geom_point(data = full_observed %>% filter(time <= fit_period), 
             aes(x = time, y = observed, color = "Observed (fitted)"), alpha = 0.5) +
  geom_point(data = full_observed %>% filter(time > fit_period), 
             aes(x = time, y = observed, color = "Observed (holdout)"), alpha = 0.5) +
  geom_vline(xintercept = fit_period, linetype = "dotted", size = 0.8) +
  annotate("text", x = fit_period - 5, y = max(forecast_df$upper_95) * 0.9, 
           label = "Fitted period", angle = 90, vjust = -0.5) +
  labs(
    title = "Epidemic Forecast with Posterior Uncertainty",
    subtitle = "Model fitted on days 0-60, forecasting to day 100",
    x = "Time (days)",
    y = "Infected individuals",
    color = NULL
  ) +
  scale_color_manual(
    values = c(
      "Forecast median" = "blue",
      "True trajectory" = "darkgreen",
      "Observed (fitted)" = "red",
      "Observed (holdout)" = "orange"
    )
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Key Observations:**

1. The forecast (blue) matches the true trajectory (green dashed) well
2. Uncertainty (shaded regions) increases as we forecast further into the future
3. The holdout observations (orange) fall within our predictive intervals
4. This demonstrates that our Bayesian model has successfully learned the epidemic dynamics!

# Summary and Key Takeaways

## What We've Learned

1. **Bayesian Workflow**: 
   - Specify a probabilistic model (priors + likelihood)
   - Fit using MCMC (Stan's HMC algorithm)
   - Check convergence (trace plots, Rhat)
   - Validate with posterior predictive checks

2. **SIR Model Fitting**:
   - Successfully recovered true parameter values from noisy data
   - Quantified uncertainty in parameter estimates
   - Made probabilistic forecasts with credible intervals

3. **Advantages of Bayesian Approach**:
   - Full posterior distributions (not just point estimates)
   - Principled uncertainty quantification
   - Flexible framework for complex models
   - Natural handling of hierarchical structures and latent variables

## Further Extensions

You could extend this analysis by:

- Adding time-varying transmission rates ($\beta(t)$) to model interventions
- Including demographic structure (age groups, spatial structure)
- Fitting to multiple data streams (deaths, hospitalizations)
- Comparing different epidemic models (SEIR, SEIRS, etc.)
- Estimating initial conditions (if $I_0$ is unknown)

## References

- Stan Development Team. "Stan Modeling Language User's Guide." https://mc-stan.org/
- Gelman et al. "Bayesian Data Analysis, Third Edition." 2013.
- Keeling & Rohani. "Modeling Infectious Diseases in Humans and Animals." 2008.

---

**Happy Bayesian modeling!** 🎉

