---
title: "Bayesian Inference for SIR Models using BayesianTools"
author: "Introduction to Bayesian Statistics"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Bayesian Inference for SIR Models using BayesianTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

This vignette demonstrates Bayesian inference for **SIR (Susceptible-Infected-Recovered) epidemic models** using the **BayesianTools** package. Unlike Stan, BayesianTools works entirely in R without requiring C++ compilation, making it more accessible for beginners.

We'll cover:

1. Understanding the SIR model (same as before!)
2. Simulating epidemic data
3. Defining likelihood and prior functions in R
4. Fitting with various MCMC algorithms (Metropolis, DEzs, DREAM)
5. Trace plots for convergence diagnostics
6. Posterior predictive checks
7. Comparing different MCMC samplers

## Why BayesianTools?

**Advantages:**

- ✅ Pure R implementation - no compilation needed
- ✅ Multiple MCMC algorithms available
- ✅ Easy to understand likelihood and prior functions
- ✅ Great for teaching Bayesian fundamentals
- ✅ Built-in diagnostic tools

**Trade-offs:**

- Slower than Stan for complex models (but fine for SIR!)
- Less automatic tuning than Stan's HMC
- May require more manual tuning for difficult posteriors

## The SIR Model (Quick Review)

The SIR model divides a population into three compartments:

$$
\begin{aligned}
\frac{dS}{dt} &= -\beta \frac{SI}{N} \\
\frac{dI}{dt} &= \beta \frac{SI}{N} - \gamma I \\
\frac{dR}{dt} &= \gamma I
\end{aligned}
$$

Parameters:

- $\beta$: transmission rate
- $\gamma$: recovery rate (1/infectious period)
- $R_0 = \beta/\gamma$: basic reproductive number

# Load Required Libraries

```{r libraries}
library(BayesianTools)
library(deSolve)
library(ggplot2)
library(dplyr)
library(tidyr)
library(coda)

# Set random seed for reproducibility
set.seed(42)
```

# Step 1: Simulate SIR Epidemic Data

We'll use the same simulation setup as the Stan vignette for comparison.

```{r simulate_sir}
# Define the SIR model as ODEs
sir_ode <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    N <- S + I + R
    
    dS <- -beta * S * I / N
    dI <- beta * S * I / N - gamma * I
    dR <- gamma * I
    
    return(list(c(dS, dI, dR)))
  })
}

# True parameter values (what we'll try to recover)
true_beta <- 0.4    # Transmission rate
true_gamma <- 0.1   # Recovery rate
true_R0 <- true_beta / true_gamma

cat("True parameters:\n")
cat(sprintf("  β = %.2f\n", true_beta))
cat(sprintf("  γ = %.2f\n", true_gamma))
cat(sprintf("  R₀ = %.2f\n", true_R0))

# Initial conditions
N <- 1000
initial_infected <- 10
initial_state <- c(
  S = N - initial_infected,
  I = initial_infected,
  R = 0
)

# Simulate the epidemic
times <- seq(0, 100, by = 1)
parameters <- c(beta = true_beta, gamma = true_gamma)

sir_solution <- ode(
  y = initial_state,
  times = times,
  func = sir_ode,
  parms = parameters
)

sir_df <- as.data.frame(sir_solution)
colnames(sir_df) <- c("time", "S", "I", "R")

# Add observation noise (Poisson)
sir_df$I_observed <- rpois(nrow(sir_df), lambda = sir_df$I)

# Plot the simulated data
ggplot(sir_df, aes(x = time)) +
  geom_line(aes(y = S, color = "Susceptible"), size = 1) +
  geom_line(aes(y = I, color = "Infected (true)"), size = 1) +
  geom_line(aes(y = R, color = "Recovered"), size = 1) +
  geom_point(aes(y = I_observed, color = "Infected (observed)"), alpha = 0.5) +
  labs(
    title = "Simulated SIR Epidemic",
    subtitle = paste0("β = ", true_beta, ", γ = ", true_gamma, ", R₀ = ", round(true_R0, 2)),
    x = "Time (days)",
    y = "Number of individuals",
    color = "Compartment"
  ) +
  theme_minimal() +
  scale_color_manual(
    values = c(
      "Susceptible" = "blue",
      "Infected (true)" = "red",
      "Infected (observed)" = "darkred",
      "Recovered" = "green"
    )
  )
```

# Step 2: Prepare Data for Fitting

```{r prepare_data}
# Use data up to day 60 for fitting
fit_period <- 60
fit_data <- sir_df %>% filter(time <= fit_period)

# Extract observed infected counts
observed_I <- fit_data$I_observed
times_fit <- fit_data$time

cat("Fitting to", length(observed_I), "observations\n")
cat("Observed infected counts (first 10 days):\n")
print(head(observed_I, 10))
```

# Step 3: Define the Bayesian Model Components

In BayesianTools, we need to define:

1. **Likelihood function**: How likely are the data given parameters?
2. **Prior function**: What do we believe about parameters before seeing data?

## 3.1 Create the Likelihood Function

```{r likelihood}
# Function to solve SIR ODE and extract infected counts
predict_sir <- function(params) {
  beta <- params[1]
  gamma <- params[2]
  
  # Solve ODE with error handling
  sir_pred <- tryCatch({
    ode(
      y = initial_state,
      times = times_fit,
      func = sir_ode,
      parms = c(beta = beta, gamma = gamma)
    )
  }, error = function(e) {
    return(NULL)
  })
  
  # Check if ODE solving failed
  if (is.null(sir_pred)) {
    return(NULL)
  }
  
  # Extract infected counts (column 3 is I)
  I_pred <- sir_pred[, 3]
  
  # Ensure non-negative predictions
  I_pred[I_pred < 0] <- 0
  
  return(I_pred)
}

# Likelihood function
# We assume observed counts follow Poisson(predicted counts)
likelihood <- function(params) {
  # Bounds checking (return very low likelihood if out of bounds)
  if (params[1] <= 0 || params[1] > 2 || 
      params[2] <= 0 || params[2] > 1) {
    return(-Inf)
  }
  
  # Get predicted infected counts
  I_pred <- predict_sir(params)
  
  # Check if ODE solving failed
  if (is.null(I_pred)) {
    return(-Inf)
  }
  
  # Check for any invalid values
  if (any(is.na(I_pred)) || any(is.infinite(I_pred))) {
    return(-Inf)
  }
  
  # Calculate log-likelihood (sum of Poisson log-likelihoods)
  # Add small constant to avoid lambda = 0
  log_lik <- sum(dpois(observed_I, lambda = pmax(I_pred, 0.01), log = TRUE))
  
  # Check if likelihood is valid
  if (is.na(log_lik) || is.infinite(log_lik)) {
    return(-Inf)
  }
  
  return(log_lik)
}

# Test the likelihood function
test_params <- c(0.4, 0.1)
cat("Log-likelihood at true parameters:", likelihood(test_params), "\n")
```

**Understanding the Likelihood:**

- For each time point, we compare observed vs. predicted infected counts
- `dpois(observed, lambda = predicted)` gives the probability of seeing the observed count if the true mean is `predicted`
- We sum log-probabilities across all time points
- Higher likelihood = better fit to data
- **Error handling**: We added `tryCatch()` and checks because the ODE solver can fail with extreme parameter values during MCMC exploration. When this happens, we return `-Inf` likelihood, which tells the sampler to reject those parameters.

## 3.2 Define Prior Distributions

```{r prior}
# Prior function
# We'll use weakly informative priors
prior_density <- function(params) {
  beta <- params[1]
  gamma <- params[2]
  
  # Check bounds - return -Inf if out of bounds
  if (beta <= 0 || beta > 2 || gamma <= 0 || gamma > 1) {
    return(-Inf)
  }
  
  # Beta ~ Normal(0.4, 0.2), truncated to (0, 2)
  log_prior_beta <- dnorm(beta, mean = 0.4, sd = 0.2, log = TRUE)
  
  # Gamma ~ Normal(0.1, 0.05), truncated to (0, 1)
  log_prior_gamma <- dnorm(gamma, mean = 0.1, sd = 0.05, log = TRUE)
  
  # Combined prior
  log_prior <- log_prior_beta + log_prior_gamma
  
  # Check for invalid values
  if (is.na(log_prior) || is.nan(log_prior)) {
    return(-Inf)
  }
  
  return(log_prior)
}

# Prior sampling function (needed for some algorithms)
prior_sampler <- function(n = 1) {
  beta <- rnorm(n, mean = 0.4, sd = 0.2)
  gamma <- rnorm(n, mean = 0.1, sd = 0.05)
  
  # Ensure positive values and within bounds
  beta <- pmax(pmin(abs(beta), 2), 0.01)
  gamma <- pmax(pmin(abs(gamma), 1), 0.01)
  
  return(cbind(beta, gamma))
}

# Test prior
test_prior <- prior_sampler(5)
cat("Sample from prior:\n")
print(test_prior)

# Test prior density
cat("\nLog prior density at true parameters:", prior_density(c(0.4, 0.1)), "\n")
```

**Understanding the Priors:**

- We defined informative priors centered around plausible values
- `β ~ Normal(0.4, 0.2)`: We expect transmission rate around 0.4
- `γ ~ Normal(0.1, 0.05)`: We expect recovery rate around 0.1 (10-day infectious period)
- These are "weakly informative" - they guide the sampler but don't overwhelm the data
- **Note**: In the setup below, we'll use uniform priors for simplicity (common in teaching examples)

## 3.3 Create the Bayesian Setup

```{r bayesian_setup}
# Create the Bayesian setup object
# Note: We can either use explicit priors OR bounds, but not both simultaneously

# OPTION 1: Use bounds (creates uniform prior) - SIMPLER, RECOMMENDED FOR TEACHING
bayesian_setup <- createBayesianSetup(
  likelihood = likelihood,
  lower = c(0.01, 0.01),      # Lower bounds (slightly above 0)
  upper = c(2, 1),            # Upper bounds
  names = c("beta", "gamma")
)

cat("Bayesian setup created successfully!\n")
cat("Using uniform priors: beta ~ Uniform(0.01, 2), gamma ~ Uniform(0.01, 1)\n")

# OPTION 2: Use informative priors (more advanced - uncomment to try)
# This requires creating a custom prior object
# prior_obj <- createPrior(
#   density = prior_density,
#   sampler = prior_sampler,
#   lower = c(0.01, 0.01),
#   upper = c(2, 1)
# )
# bayesian_setup <- createBayesianSetup(
#   likelihood = likelihood,
#   prior = prior_obj,
#   names = c("beta", "gamma")
# )
```

# Step 4: Fit the Model Using Different MCMC Algorithms

BayesianTools offers several MCMC algorithms. We'll try three:

1. **DEzs (Differential Evolution with snooker updater)**: Efficient for many problems
2. **DREAM (DiffeRential Evolution Adaptive Metropolis)**: Good for multimodal posteriors
3. **Metropolis**: Classic algorithm (for comparison)

## 4.1 Fit with DEzs (Recommended)

```{r fit_dezs}
cat("Fitting with DEzs algorithm...\n")

fit_dezs <- runMCMC(
  bayesianSetup = bayesian_setup,
  sampler = "DEzs",
  settings = list(
    iterations = 10000,   # Number of iterations
    nrChains = 3,         # Number of chains
    burnin = 2000,        # Burn-in period
    thin = 5,             # Thinning interval
    message = TRUE
  )
)

cat("\nDEzs fitting complete!\n")
```

## 4.2 Fit with DREAM

```{r fit_dream}
cat("Fitting with DREAM algorithm...\n")

fit_dream <- runMCMC(
  bayesianSetup = bayesian_setup,
  sampler = "DREAM",
  settings = list(
    iterations = 10000,
    nrChains = 3,
    burnin = 2000,
    thin = 5,
    message = TRUE
  )
)

cat("\nDREAM fitting complete!\n")
```

## 4.3 Fit with Metropolis

```{r fit_metropolis}
cat("Fitting with Metropolis algorithm...\n")

fit_metropolis <- runMCMC(
  bayesianSetup = bayesian_setup,
  sampler = "Metropolis",
  settings = list(
    iterations = 50000,   # More iterations needed for Metropolis
    nrChains = 3,
    burnin = 10000,
    adaptationNotBefore = 20000,
    thin = 10,
    message = TRUE,
    startValue = c(0.4, 0.1)  # Starting values
  )
)

cat("\nMetropolis fitting complete!\n")
```

# Step 5: Examine Results

## 5.1 Summary Statistics

```{r summary}
cat("========================================\n")
cat("SUMMARY: DEzs Algorithm\n")
cat("========================================\n")
summary(fit_dezs)

cat("\n========================================\n")
cat("SUMMARY: DREAM Algorithm\n")
cat("========================================\n")
summary(fit_dream)

cat("\n========================================\n")
cat("SUMMARY: Metropolis Algorithm\n")
cat("========================================\n")
summary(fit_metropolis)

# Compare to true values
cat("\n========================================\n")
cat("COMPARISON TO TRUE VALUES\n")
cat("========================================\n")

# Extract posterior means from DEzs (best performing)

posterior_means <- getSample(fit_dezs, start = 2, thin = 1) %>% apply(2, mean)


cat(sprintf("Beta:  True = %.3f, Estimated (DEzs) = %.3f\n", 
            true_beta, posterior_means[1]))
cat(sprintf("Gamma: True = %.3f, Estimated (DEzs) = %.3f\n", 
            true_gamma, posterior_means[2]))
cat(sprintf("R0:    True = %.3f, Estimated (DEzs) = %.3f\n", 
            true_R0, posterior_means[1]/posterior_means[2]))
            
```

## 5.2 Convergence Diagnostics

```{r convergence}
# Gelman-Rubin diagnostic (Rhat)
# Values close to 1.0 indicate convergence
# Rhat < 1.1 is generally considered good

cat("Gelman-Rubin Diagnostic (Rhat):\n")
cat("Values < 1.1 indicate good convergence\n\n")

cat("DEzs:\n")
gelmanDiagnostics(fit_dezs)

cat("\nDREAM:\n")
gelmanDiagnostics(fit_dream)

cat("\nMetropolis:\n")
gelmanDiagnostics(fit_metropolis)
```

# Step 6: Trace Plots

Trace plots show parameter values across MCMC iterations. Good mixing looks like a "fuzzy caterpillar."

## 6.1 DEzs Trace Plots

```{r trace_dezs, fig.height=8}
plot(fit_dezs, which = c(1, 2))
tracePlot(fit_dezs, start = 2)
```

**What to look for:**

- ✅ All chains (colors) overlap and explore same region
- ✅ No trends or drifts over time
- ✅ "Fuzzy caterpillar" appearance
- ✅ Good mixing between chains

## 6.2 DREAM Trace Plots

```{r trace_dream, fig.height=8}
plot(fit_dream, which = c(1, 2))
tracePlot(fit_dream, start = 2)
```

## 6.3 Metropolis Trace Plots

```{r trace_metropolis, fig.height=8}

plot(fit_metropolis, which = c(1, 2))
tracePlot(fit_metropolis, start = 1)

```

**Comparison:**

- DEzs and DREAM typically show better mixing than basic Metropolis
- Metropolis may show autocorrelation (slower exploration)
- All three should converge to the same posterior if run long enough

# Step 7: Posterior Distributions

## 7.1 Marginal Posteriors

```{r marginal_posteriors}
# Plot marginal posterior distributions
par(mfrow = c(2, 2))

# DEzs
plot(fit_dezs, which = c(3, 4))
abline(v = true_beta, col = "red", lwd = 2, lty = 2)
abline(v = true_gamma, col = "red", lwd = 2, lty = 2)
```

## 7.2 Joint Posterior (Correlation Plot)

```{r correlation_plot}
# Correlation between beta and gamma
correlationPlot(fit_dezs)
```

**Interpretation:**

- Shows the joint posterior distribution of β and γ
- Reveals any correlations between parameters
- Elliptical shape indicates correlation
- Red lines show true parameter values

## 7.3 Posterior for R0

```{r r0_posterior}
# Calculate R0 from posterior samples
samples_dezs <- getSample(fit_dezs, start = 2, thin = 1)
R0_posterior <- samples_dezs[, 1] / samples_dezs[, 2]

# Plot R0 posterior
hist(R0_posterior, breaks = 50, 
     main = "Posterior Distribution of R₀",
     xlab = "R₀",
     col = "skyblue",
     border = "white",
     freq = FALSE)
abline(v = true_R0, col = "red", lwd = 3, lty = 2)
abline(v = mean(R0_posterior), col = "blue", lwd = 2)
legend("topright", 
       legend = c("True R₀", "Posterior mean"),
       col = c("red", "blue"),
       lty = c(2, 1),
       lwd = c(3, 2))

# Summary statistics for R0
cat("\nPosterior summary for R₀:\n")
cat(sprintf("  Mean: %.2f\n", mean(R0_posterior)))
cat(sprintf("  Median: %.2f\n", median(R0_posterior)))
cat(sprintf("  95%% CI: [%.2f, %.2f]\n", 
            quantile(R0_posterior, 0.025),
            quantile(R0_posterior, 0.975)))
cat(sprintf("  True value: %.2f\n", true_R0))
```

# Step 8: Posterior Predictive Checks

Let's check if our model can reproduce the observed data.

```{r posterior_predictive}
# Extract posterior samples (use DEzs as best performer)
n_samples <- 500
sample_idx <- sample(nrow(samples_dezs), n_samples)
posterior_samples <- samples_dezs[sample_idx, ]

# Generate predictions for each posterior sample
predictions <- matrix(NA, nrow = n_samples, ncol = length(times_fit))

for (i in 1:n_samples) {
  predictions[i, ] <- predict_sir(as.numeric(posterior_samples[i, ]))
}

# Calculate quantiles
pred_quantiles <- apply(predictions, 2, quantile, 
                        probs = c(0.025, 0.1, 0.5, 0.9, 0.975))

# Create data frame for plotting
pred_df <- data.frame(
  time = times_fit,
  observed = observed_I,
  median = pred_quantiles[3, ],
  lower_95 = pred_quantiles[1, ],
  upper_95 = pred_quantiles[5, ],
  lower_80 = pred_quantiles[2, ],
  upper_80 = pred_quantiles[4, ]
)

# Plot posterior predictive distribution
ggplot(pred_df, aes(x = time)) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), 
              fill = "skyblue", alpha = 0.3) +
  geom_ribbon(aes(ymin = lower_80, ymax = upper_80), 
              fill = "skyblue", alpha = 0.5) +
  geom_line(aes(y = median, color = "Predicted median"), size = 1) +
  geom_point(aes(y = observed, color = "Observed data"), alpha = 0.6) +
  labs(
    title = "Posterior Predictive Check",
    subtitle = "Model predictions vs. observed data",
    x = "Time (days)",
    y = "Infected individuals",
    color = NULL
  ) +
  scale_color_manual(values = c("Predicted median" = "blue", 
                                 "Observed data" = "red")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Interpretation:**

- **Shaded regions**: Uncertainty in predictions (darker = 80% CI, lighter = 95% CI)
- **Blue line**: Median prediction
- **Red points**: Observed data

If the model fits well, most observed points should fall within the predictive intervals. ✓

# Step 9: Forecasting

Now let's forecast the epidemic beyond day 60 using our fitted model.

```{r forecast}
# Forecast time points (full 100 days)
times_forecast <- seq(0, 100, by = 1)

# Generate forecasts
n_forecast <- 500
forecast_idx <- sample(nrow(samples_dezs), n_forecast)
forecast_samples <- samples_dezs[forecast_idx, ]

# Store forecasts
forecast_I <- matrix(NA, nrow = n_forecast, ncol = length(times_forecast))

for (i in 1:n_forecast) {
  beta_i <- as.numeric(forecast_samples[i, 1])
  gamma_i <- as.numeric(forecast_samples[i, 2])
  
  sir_forecast <- ode(
    y = initial_state,
    times = times_forecast,
    func = sir_ode,
    parms = c(beta = beta_i, gamma = gamma_i)
  )
  
  forecast_I[i, ] <- sir_forecast[, 3]
}

# Calculate quantiles
forecast_quantiles <- apply(forecast_I, 2, quantile, 
                            probs = c(0.025, 0.1, 0.5, 0.9, 0.975))

forecast_df <- data.frame(
  time = times_forecast,
  median = forecast_quantiles[3, ],
  lower_95 = forecast_quantiles[1, ],
  upper_95 = forecast_quantiles[5, ],
  lower_80 = forecast_quantiles[2, ],
  upper_80 = forecast_quantiles[4, ]
)

# Add observed data
full_observed <- data.frame(
  time = sir_df$time,
  observed = sir_df$I_observed,
  true = sir_df$I
)

# Plot forecast
ggplot() +
  geom_ribbon(data = forecast_df, 
              aes(x = time, ymin = lower_95, ymax = upper_95), 
              fill = "skyblue", alpha = 0.3) +
  geom_ribbon(data = forecast_df, 
              aes(x = time, ymin = lower_80, ymax = upper_80), 
              fill = "skyblue", alpha = 0.5) +
  geom_line(data = forecast_df, 
            aes(x = time, y = median, color = "Forecast median"), 
            size = 1) +
  geom_line(data = full_observed, 
            aes(x = time, y = true, color = "True trajectory"), 
            linetype = "dashed", size = 1) +
  geom_point(data = full_observed %>% filter(time <= fit_period), 
             aes(x = time, y = observed, color = "Observed (fitted)"), 
             alpha = 0.5) +
  geom_point(data = full_observed %>% filter(time > fit_period), 
             aes(x = time, y = observed, color = "Observed (holdout)"), 
             alpha = 0.5) +
  geom_vline(xintercept = fit_period, linetype = "dotted", size = 0.8) +
  annotate("text", x = fit_period - 5, y = max(forecast_df$upper_95) * 0.9, 
           label = "Fitted period", angle = 90, vjust = -0.5) +
  labs(
    title = "Epidemic Forecast with Posterior Uncertainty",
    subtitle = "Model fitted on days 0-60, forecasting to day 100",
    x = "Time (days)",
    y = "Infected individuals",
    color = NULL
  ) +
  scale_color_manual(
    values = c(
      "Forecast median" = "blue",
      "True trajectory" = "darkgreen",
      "Observed (fitted)" = "red",
      "Observed (holdout)" = "orange"
    )
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Key Observations:**

1. ✓ Forecast matches true trajectory well
2. ✓ Uncertainty increases further into future
3. ✓ Holdout observations fall within predictive intervals
4. ✓ Model has learned the epidemic dynamics!

# Step 10: Compare MCMC Algorithms

Let's compare the efficiency of different samplers.

```{r compare_samplers}
# Extract effective sample sizes
cat("Effective Sample Sizes:\n")
cat("(Higher is better - more independent samples)\n\n")

cat("DEzs:\n")
effectiveSize(getSample(fit_dezs, start = 2, thin = 1))

cat("\nDREAM:\n")
effectiveSize(getSample(fit_dream, start = 2, thin = 1))

cat("\nMetropolis:\n")
effectiveSize(getSample(fit_metropolis, start = 2, thin = 1))

# Create comparison plot
samples_list <- list(
  DEzs = getSample(fit_dezs, start = 2, thin = 1),
  DREAM = getSample(fit_dream, start = 2, thin = 1),
  Metropolis = getSample(fit_metropolis, start = 2, thin = 1)
)

# Plot posterior densities from all three methods
par(mfrow = c(1, 2))

# Beta
plot(density(samples_list$DEzs[, 1]), 
     main = "Posterior: β (transmission rate)",
     xlab = "β", col = "blue", lwd = 2, xlim = c(0.3, 0.5))
lines(density(samples_list$DREAM[, 1]), col = "green", lwd = 2)
lines(density(samples_list$Metropolis[, 1]), col = "red", lwd = 2)
abline(v = true_beta, lty = 2, lwd = 2)
legend("topright", 
       legend = c("DEzs", "DREAM", "Metropolis", "True"),
       col = c("blue", "green", "red", "black"),
       lty = c(1, 1, 1, 2),
       lwd = 2)

# Gamma
plot(density(samples_list$DEzs[, 2]), 
     main = "Posterior: γ (recovery rate)",
     xlab = "γ", col = "blue", lwd = 2, xlim = c(0.08, 0.12))
lines(density(samples_list$DREAM[, 2]), col = "green", lwd = 2)
lines(density(samples_list$Metropolis[, 2]), col = "red", lwd = 2)
abline(v = true_gamma, lty = 2, lwd = 2)
legend("topright", 
       legend = c("DEzs", "DREAM", "Metropolis", "True"),
       col = c("blue", "green", "red", "black"),
       lty = c(1, 1, 1, 2),
       lwd = 2)

par(mfrow = c(1, 1))
```

**Sampler Comparison:**

- **DEzs & DREAM**: Modern algorithms with good efficiency
- **Metropolis**: Classic algorithm, may need more iterations
- All three converge to similar posteriors (good sign!)

# Summary and Key Takeaways

## What We've Learned

### 1. BayesianTools Workflow

1. Define likelihood function (data generating process)
2. Specify prior distributions
3. Create Bayesian setup
4. Run MCMC sampler
5. Check convergence
6. Validate with posterior predictive checks

### 2. MCMC Algorithms

- **DEzs**: Good default choice, efficient for many problems
- **DREAM**: Excellent for complex posteriors
- **Metropolis**: Simple but may need more iterations

### 3. Key Diagnostics

- **Rhat < 1.1**: Good convergence
- **Effective sample size**: More is better
- **Trace plots**: Should look like "fuzzy caterpillars"
- **Posterior predictive checks**: Model should reproduce data

## BayesianTools vs. Stan

| Feature | BayesianTools | Stan |
|---------|--------------|------|
| Setup | Pure R, no compilation | Requires C++ compilation |
| Speed | Moderate | Fast (for complex models) |
| Learning curve | Gentle | Steeper |
| Flexibility | Good | Excellent |
| Diagnostics | Built-in tools | Excellent (via bayesplot) |
| Best for | Teaching, simple-moderate models | Complex hierarchical models |

**Bottom line**: Use BayesianTools for learning and moderate-complexity problems. Graduate to Stan for production work and complex models.

## Further Extensions

You could extend this analysis by:

- Trying different prior distributions
- Adding measurement error models
- Including time-varying parameters
- Fitting SEIR or other compartmental models
- Using adaptive MCMC (`sampler = "AM"`)
- Implementing model comparison with DIC or WAIC

## References

- Hartig, F. (2022). BayesianTools: General-Purpose MCMC and SMC Samplers and Tools for Bayesian Statistics. R package.
- Gelman et al. (2013). *Bayesian Data Analysis, Third Edition*. CRC Press.
- ter Braak & Vrugt (2008). Differential Evolution Markov Chain with snooker updater. *Statistics and Computing*.

---

**Happy Bayesian modeling with BayesianTools!** 🎉

